---
output:
  github_document:
    html_preview: false
params:
  site_counts: NA
  global_grid: NA
  yaml_contents: NA
always_allow_html: true
---

# Download process

```{r message=FALSE, warning = FALSE, echo=FALSE}
global_grid <- params$global_grid
site_counts <- params$site_counts
yaml_contents <- params$yaml_contents

options(tigris_use_cache = TRUE)
```

### Catalogue existing data

The download process begins by using the {dataRetrieval} package to catalogue the data that is available in the WQP. To do this the user selects parameters of interest and their corresponding `characteristicName` for retrieving data from the WQP. Below are the current parameters and their `characteristicName`s as defined in the configuration YAML, `1_inventory/cfg/wqp_codes.yml`. Data for these parameters are requested from the WQP within a spatial grid that is mapped in the next section.

```{r, results='asis'}
names_yaml <- names(yaml_contents)

map_df(.x = names_yaml,
    .f = ~yaml_contents[.x] %>%
        as_tibble() %>%
        rename(characteristicName = 1) %>%
        mutate(parameter = .x) %>%
        select(parameter, characteristicName)) %>%
  kable() %>%
  kable_material() %>%
  collapse_rows(columns = 1, valign = "top")
```

<br>

### Maps of data spread

The pipeline uses a spatial grid to download data from the WQP in batches. Maps are presented below with counts of records across this spatial grid. **Note:** The counts here are for raw data that are not yet filtered or harmonized.

```{r}
# Combine sample counts in each grid_id with the grid polygons
grid_counts <- left_join(x = global_grid,
                         y = site_counts %>%
                           count(grid_id),
                         by = c("id" = "grid_id")) %>%
  st_transform(crs = 9311)

state_selection <- states(progress_bar = FALSE) %>%
  st_transform(crs = 9311)

# Conterminous US map:
conterminous_us <- state_selection %>%
  filter(!(NAME %in% c("Alaska", "Hawaii", "American Samoa",
                       "Guam", "Puerto Rico",
                       "United States Virgin Islands",
                       "Commonwealth of the Northern Mariana Islands")))

ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = conterminous_us,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(min(st_coordinates(conterminous_us)[,"X"]),
                    max(st_coordinates(conterminous_us)[,"X"])),
           ylim = c(min(st_coordinates(conterminous_us)[,"Y"]),
                    max(st_coordinates(conterminous_us)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("Conterminous US") +
  theme_bw()

# Alaska map:
AK <- state_selection %>%
  filter(NAME == "Alaska")

ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = AK,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(min(st_coordinates(AK)[,"X"]),
                    max(st_coordinates(AK)[,"X"])),
           ylim = c(min(st_coordinates(AK)[,"Y"]),
                    max(st_coordinates(AK)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("Alaska") +
  theme_bw()

# Hawaii map:
HI <- state_selection %>%
  filter(NAME == "Hawaii")

ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = HI,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(min(st_coordinates(HI)[,"X"]),
                    0.9 * max(st_coordinates(HI)[,"X"])),
           ylim = c(1.1 * min(st_coordinates(HI)[,"Y"]),
                    max(st_coordinates(HI)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("Hawaii") +
  theme_bw()


# American Samoa map:
AS <- state_selection %>%
  filter(NAME %in% c("American Samoa"))


ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = AS,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(1.025 * min(st_coordinates(AS)[,"X"]),
                    0.975 *max(st_coordinates(AS)[,"X"])),
           ylim = c(1.05 * min(st_coordinates(AS)[,"Y"]),
                    max(st_coordinates(AS)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("American Samoa") +
  theme_bw()


# Guam & Commonwealth of the Northern Mariana Islands map
GU_CNMI <- state_selection %>%
  filter(NAME %in% c("Guam", "Commonwealth of the Northern Mariana Islands"))

ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = GU_CNMI,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(1.025 * min(st_coordinates(GU_CNMI)[,"X"]),
                    0.975 * max(st_coordinates(GU_CNMI)[,"X"])),
           ylim = c(0.95 * min(st_coordinates(GU_CNMI)[,"Y"]),
                    1.05 * max(st_coordinates(GU_CNMI)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("Guam & Commonwealth of the Northern Mariana Islands") +
  theme_bw()


# Puerto Rico & United States Virgin Islands map:
PR_VI <- state_selection %>%
  filter(NAME %in% c("Puerto Rico", "United States Virgin Islands"))


ggplot() +
  geom_sf(data = grid_counts,
          aes(fill = n)) +
  geom_sf(data = PR_VI,
          fill = NA, color = "black") +
  xlab(NULL) +
  ylab(NULL) +
  coord_sf(xlim = c(0.95 * min(st_coordinates(PR_VI)[,"X"]),
                    1.05 * max(st_coordinates(PR_VI)[,"X"])),
           ylim = c(1.075 * min(st_coordinates(PR_VI)[,"Y"]),
                    0.925 * max(st_coordinates(PR_VI)[,"Y"]))) +
  scale_fill_viridis_c("Record count",
                       trans = "log10",
                       labels = scales::label_number(),
                       na.value = "white",
                       breaks = c(1, 10, 100, 1000, 10000)) +
  ggtitle("Puerto Rico & United States Virgin Islands") +
  theme_bw()

```
