---
output:
  github_document:
    html_preview: false
params:
  documented_drops: NA
---

# Pre-harmonization decision process

```{r echo=FALSE}
documented_drops <- params$documented_drops
```

```{r echo = FALSE}
step_order_rev <- max(documented_drops$order):min(documented_drops$order)

step_data <- documented_drops %>%
  group_by(step) %>%
  mutate(scaled = scale(n_rows, center = FALSE)) %>%
  ungroup() %>%
  mutate(order = factor(order,
                        levels = step_order_rev),
         rows_label = paste0(short_reason,
                             ":  ",
                             round((n_rows / 1000000), 1),
                             " M rows")) %>%
  filter(step == "pre-harmonization") 
```

### Initial dataset

```{r echo=FALSE}
step_0_rows <- filter(step_data,
                      step == "pre-harmonization",
                      order == 0) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")
```

At the start of the preharmonization process the WQP dataset contains `r step_0_rows` rows.

```{r echo = FALSE}
step_data %>%
  filter(order == 0) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

<br>

### Remove duplicates

```{r echo = FALSE}
step_1_rows <- filter(step_data,
                      order == 1) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")

step_1_dropped <- filter(step_data,
                         order == 1) %>%
  mutate(mil_row = number_dropped / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")
```

The first filtering step is to remove any duplicated records present in the dataset. Duplicate records occur in the WQP database. The pipeline allows for the user to specify which columns to use when checking for duplicated information. WQP columns currently used to check for duplicates include: OrganizationIdentifier, MonitoringLocationIdentifier, ActivityStartDate, ActivityStartTime.Time, CharacteristicName, and ResultSampleFractionText. This results in dropping `r step_1_dropped` rows for a record count of `r step_1_rows`.

```{r echo = FALSE}
step_data %>%
  filter(order %in% 0:1) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[2:1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

<br>

### Missing results

```{r echo = FALSE}
step_2_rows <- filter(step_data,
                      order == 2) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")

step_2_dropped <- filter(step_data,
                         order == 2) %>%
  mutate(thou_row = number_dropped / 1000) %>%
  pull(thou_row) %>%
  round(digits = 2) %>%
  paste0(., " thousand")
```

Next records that have missing data are dropped from the dataset. Several criteria are used when checking for missing data. If any of the below criteria are met the row is flagged as missing:

1. Both the result column and detection limit column had `NA` data
2. Result, result unit, activity comment, laboratory comment, and result comment columns are all `NA`
3. Result detection condition column contains `not reported`
4. The result comment column contains any user-provided text indicating a missing value, currently including: `analysis lost`, `not analyzed`, `not recorded`, `not collected`, or `no measurement taken`

`r step_2_dropped` rows are dropped, resulting in a final count of `r step_2_rows`.


```{r echo = FALSE}
step_data %>%
  filter(order %in% 0:2) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[3:1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

<br>

### Filter status

```{r echo = FALSE}
step_3_rows <- filter(step_data,
                      order == 3) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")

step_3_dropped <- filter(step_data,
                         order == 3) %>%
  mutate(thou_row = number_dropped / 1000) %>%
  pull(thou_row) %>%
  round(digits = 2) %>%
  paste0(., " thousand")
```

We next filter the `ResultStatusIdentifier` column to include only the following statuses:

+ `"Accepted"` 
+ `"Final"`
+ `"Historical"`
+ `"Validated"`
+ `"Preliminary"`
+ `NA`

These statuses generally indicate a reliable result having been reached, however we also include `NA` in an effort to be conservative. More specifically, when making decisions for this and other columns we occasionally retain `NA` values if removing the records would otherwise drop 10% or more of the available data. `r step_3_dropped` rows are dropped from the dataset leaving it with `r step_3_rows` remaining.

```{r echo = FALSE}
step_data %>%
  filter(order %in% 0:3) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[4:1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

<br>

### Filter media

```{r echo = FALSE}
step_4_rows <- filter(step_data,
                      order == 4) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")

step_4_dropped <- filter(step_data,
                         order == 4) %>%
  mutate(thou_row = number_dropped / 1000) %>%
  pull(thou_row) %>%
  round(digits = 2) %>%
  paste0(., " thousand")
```

Next the `ActivityMediaSubdivisionName` is filtered to only include only the following media types:

+ `"Surface Water"`
+ `"Water"`
+ `"Estuary"`
+ `NA`

`r step_4_dropped` rows are dropped and `r step_4_rows` remain.

```{r echo = FALSE}
step_data %>%
  filter(order %in% 0:4) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[5:1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

<br>

#### Location type

```{r echo = FALSE}
step_5_rows <- filter(step_data,
                      order == 5) %>%
  mutate(mil_row = n_rows / 1000000) %>%
  pull(mil_row) %>%
  round(digits = 2) %>%
  paste0(., " million")

step_5_dropped <- filter(step_data,
                         order == 5) %>%
  mutate(thou_row = number_dropped / 1000) %>%
  pull(thou_row) %>%
  round(digits = 2)
```

The final step in pre-harmonization is filtering out any `ResolvedMonitoringLocationTypeName` that is not `"Estuary"`, `"Lake, Reservoir, Impoundment"`, or `"Stream"`.

After dropping `r step_5_dropped` rows the pre-harmonization dataset is complete with `r step_5_rows`.

```{r echo = FALSE}
step_data %>%
  filter(order %in% 0:5) %>%
  ggplot() +
  geom_bar(aes(x = order, y = scaled, fill = order),
           stat = "identity")  +
  geom_text_repel(aes(x = order, y = 0.1, label = rows_label),
                  bg.color = "white", bg.r = 0.15,
                  point.size = NA,
                  xlim = c(-Inf, Inf),
                  ylim =  c(-Inf, Inf),
                  nudge_x = 0.1,
                  hjust = "left") +
  xlab("Step number") +
  ylab(NULL) +
  scale_fill_manual(values = viridis(n = 11, direction = -1)[6:1]) +
  scale_x_discrete(drop = F) +
  coord_flip() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```
